\begin{abstract}

Free will has been studied more than two thousand years. Past philosophers care more about how we decide on choices, as reflected by a popular formulation of free will as \emph{the ability to do otherwise}. In this paper, I'll argue that we should pay more attention to choices themselves rather than how they are decided on. In the real world, agents are firstly and mostly faced with problems, then they generate candidate choices for decision and actions. More \emph{creative} an agent is, more and better choices can be generated, decided and acted upon, thus more free the agent is. Free will is first and foremost related to \emph{creativity}.

Based on this analysis, I propose that \emph{free will is the ability to achieve an objective creatively}. According to this new conception, free will is \emph{creative}, it varies in degrees and can be improved if the agent's \emph{creativity} can be improved.

This new conception avoids the hard metaphysical problems that past philosophers have faced. Another benefit we gain from the new conception is that it's actually closer to our experience, and answers a lot of questions they feel difficult to answer, such as when a child has free will. Finally, the new conception provides a new perspective on responsibility, the insights will no doubt be very useful in education and management regarding awarding and punishment.

% Past philosophers either think \emph{free will} as too noble and commit too much to it -- for example, assume the existence of \emph{causally efficacious} and \emph{non-physical} agents, which contradicts the contemporary scientific belief about the \emph{causal closedness} of the world. Or, they think \emph{free will} as too humble, such as admitting there's no \emph{real alternatives}, which makes it difficult to support ideas like human dignity, choices and moral responsibility, which are thought to be the foundation of human society.

% The new conception gives us more things that the second approach can't provide. For example, human beings can still be proud of their free will, as human creativity is orders higher than other creatures in a lot of aspects. We still have choices as long as we can learn to be more creative, or more rational, or update our values. We can still be held as responsible if our values can be updated, or our creativity or rationality can be improved.

However, I don't claim that all problems about free will are solved, as people may ask \emph{what's the essence of human creativity}. Nevertheless, \emph{creativity} is a concept much less mysterious and less metaphysical than \emph{free will}, as we are acquainted with animals in daily life that can also exhibit creativity. What I tried to do and hopefully managed to doing in this paper is to demystify free will.

\end{abstract}
