\section{Free Will Is Creative}

The philosopher Jean-Paul Sartre has a famous quote saying that \emph{man is condemned to be free}. For Sartre and many other philosophers, free will seems to be a magic power, once owned by an agent, can be exerted on any matters of affair. This view is a thousand year old myth, which is the source of all confusions about free will.

In this chapter, I'll try to break this myth and suggest a new conception of free will that's based on the capacity to generate choices rather than the capacity to choose.

% According to this view, free will is an inexhaustible static asset for agents. I'll argue in this section that free will is not a static asset, but a dynamic asset. Free will is not universal money that can buy things everywhere, but coupons that must be used at the corresponding stores.

\subsection{The Myth of Choice}

The dominant view among philosophers is to regard free will as the capacity to choose, as reflected by the popular formulation of free will as \emph{the capacity to do otherwise}. I call this view \emph{the myth of choice}. We can see the traces of this view in all debates about free will. It's the view advocated by libertarians, but also the view that compatibilists are struggling with. Researchers studying the possibility of free will from the perspective of quantum physics and neural science are also targeting this view of free will. If we were able to break this myth, then we would fundamentally change how people will debate and talk about free will in future. This is the task I'm going to undertake here.

In the beginning of this paper, we've seen an example where you can decide freely on whether to give the coin in your pocket to the street artist or not. It seems that each choice is at your disposal, it's completely up to you to decide. For both traditional philosophers and ordinary people, this feel of ability to make free choice seems to be essential to the concept of free will.

First of all, intuition is by no means the criteria of correctness. Every human being living on earth has strong intuition that the earth is flat, but it's not true. Our intuition of free choice might just be an illusion. As it's widely known, philosophers working intentions and actions, psychologists working on subconsciousness, physiologists working on mental disorders have already greatly weakened our intuitions of free choice.

Second, free will is not irrationality. In another word, free will is rational. It means you can't choose either A or B, you should choose the better one. If you deliberately choose the suboptimal choice, we'll not say you have free will, instead we suspect if you have some mental problems.

You might ask, does rationality reflect free will? To answer this question, let's see a short program.

\begin{lstlisting}[language=Scala]
  def decide(A, B) = {
    if (A betterThan B)
      choose(A)
    else if (B betterThan A)
      choose(B)
    else
      ...
  }
\end{lstlisting}

As you can see in this program, the method \emph{decide} chooses between A and B, and it's obvious that it always chooses the better one. We have to admit the program is rational, because it always makes the better choice. Does it make sense to attribute free will to such a simple mechanism? I think this capability is too trivial to qualify the name free will.

Third, \emph{free will is not hesitation}. When you can't decide between two choices, you don't know what to do next, you feel unease about the uncertainty, you want to get out of the situation. You linger on one choice, while looking at the other. You intend to decide on one, but you hesitate. You swing like a pendulum between the two choices. This is a trouble instead of freedom!

This insight is also captured by the \emph{Paradox of Buridan's Ass}\footnote{\url{http://en.wikipedia.org/wiki/Buridan's_ass}}, which was first put forward by a 14th century French philosopher Jean Buridan. The paradox says that if there are two stacks of hay in front of a rational ass with equal distance, then the rational ass would starve to death as it can't decide which stack of hay to eat. In this case, hesitation is a synonym of being trapped, how can we say being trapped is an exhibition of free will?

From a different respect, two equally good choices might not be equal in fact, it's only that you can't tell one from the other. So in this regard, hesitation is usually a reflection of inability to decide efficiently. In such cases, you need help from experts or you need more training in the field.

If you think hesitation as some capacity that qualifies the name of free will, following program will help you to get rid of the idea.

\begin{lstlisting}[language=Scala]
  def decide(A, B) = {
    if (A betterThan B)
      choose(A)
    else if (B betterThan A)
      choose(B)
    else
      decide(B, A)
  }
\end{lstlisting}

In the program, if A and B are not equally good, then the better one is decided. If A and B are equally good, then the decide function would run forever -- that's to hesitate forever. This capacity is too trivial to qualify the name of free will.

% Some may object that the hesitation of a program is not true hesitation, because the program doesn't really understand the choices it's going to make. Now let's imagine there exists an agent, which understands the consequences of each choice, but doesn't have any desires, objectives or values at all.

Fouth, \emph{free will is not randomness}. When facing two equally good choices, we often resort to the strategy of random choice. Resorting to random choice doesn't necessarily mean resort to a true random process in the physical world. It just means to give up our control, and let external forces -- which we can't predict the result -- to decide. As we are passive and determined by external forces when we resort to randomness in decision, so the process doesn't reflect free will at all.

On the other hand, most cases of resorting to randomness in daily lives imply that you want to try your luck, which in turn implies you are aware that the two choices are most likely not equal, but you are unable to differentiate them. So most cases of resorting to randomness reflect the inability to choose. It's unreasonable to call this inability free will.

One may disagree on one point: isn't the act of resorting to randomness in case of equally good choices an exhibition of free will? Maybe, but the ability can also be implemented by a simple program:

\begin{lstlisting}[language=Scala]
def decide(A, B) = {
  if (A betterThan B)
     choose(A)
  else if (B betterThan A)
     choose(B)
  else
     choose(random(A, B))
}
\end{lstlisting}

As you can see in the program, if A and B are equally good, it would randomly choose one; otherwise, it would choose the better one.

Is it possible for any conception of free will be deviant from the program regarding the capacity to choose? I can hardly imagine any. First, the decision process should be rational to some extent, i.e. to choose the better choice, or it doesn't qualify the name of free will. The rational decision process will generate a unique result as long as there's only one best answer. Second, in case of truly equally good answers, the decision process should choose one randomly, to avoid the \emph{Paradox of Buridan's Ass}.

As I've showed above, there's nothing left in the capacity to choose that can honor free will. It's time to declare the bankruptcy of the myth of choice. It's not the capacity to choose that matters to free will, but something else.

\subsection{It's About Choices, Not Choice}

To break a myth is not enough, we have to replace it with another one. Next I'm going to argue that it's not the capability to choose but the capability to generate choices that matters.

Nearly all definitions of free will begin with the situation where the agent is about to make a choice. Some philosophers emphasize the ability to make a different choice, some emphasize the alignment between choice and action.

But making a choice is not the primitive situation we face in daily lives. In the real world, we first face a problem, then we generate candidate choices for decisions. We've seen that the capacity to choose is unable to honor free will, then the only possibility lies with the choices. The more creative we are, the better choices can be generated and decided, thus more free we are. Free will is first and foremost a problem related to \emph{creativity}. It's about choices, not choice!

From a linguistic point of view, what past philosophers care about is the adverb which modifies the verb in following sentence:

\begin{quote}
  The agent \emph{freely} decided on one among the three options.
\end{quote}

Instead, the approach I advocate in this paper would rewrite the sentence and emphasize the \emph{options} part:

\begin{quote}
  The agent decided on the best one among \emph{the three options he created}.
\end{quote}

Note that in the rewriting, \emph{freely} is removed while \emph{best} is added. This corresponds to the conclusion in previous section that the capacity to choose doesn't qualify the name of free will and the decision process should be rational to choose the best option.

If the three options are indeed created by the agent himself, the rewriting literally means the same thing. If the three options are provided by third parties, we need more complex rewriting. When we face choices provided by third parties, it's first and foremost a problem(not choice), and second a problem related to choice. We can generate different choices in order to solve the choice problem. We have to decide on one choice we generated in order to decide on the choice problem.

For example, John receives two offers from two universities, he has to decide which one to decline. He would generate two choices in order to decide on the choice problem:

\begin{quote}
(1) Choose the university which has the top overall ranking \\
(2) Choose the university which has the top specialty ranking
\end{quote}

After generating the two choices, John decided on the second one, because he's persuaded by his father that the ranking of specialty is more important than school. Then, he would be able to choose a university by the criteria he has chosen.

It's obvious that the quality of the decision mainly depends on the quality of the choices that are generated. The better choices we can generate, the greater chance we can make a better decision. If we move the focus of free will from the capacity to choose to the capacity to generate choices, we see a very different scenery. It will cast new light in the understanding of free will and reshape future debates about free will.

To make the discussion more concrete, let me first put forward a new definition of free will formally:

\begin{quote}
\emph{Free will} is the ability to creatively generate action plans and choose one rationally with respect to a specific objective in a concrete scenario.
\end{quote}

This definition doesn't sound as good as \emph{free will is creative}, but in spirit they are the same. In the following, I'll analyze this new definition in details.

\subsection{Creativity}

The definition depends on the concept \emph{action plan}, which is a list of actions to be executed in order to achieve a specific objective in a concrete scenario. When facing a problem, each action plan is a choice. Action plan is relative to the objective, each action in the action plan can also be an objective for sub-action plans. For example, suppose visit Paris is an action plan for having a good holiday. In turn, go to Paris by train is a possible step of an action plan for visiting Paris.

The interplay between action plan and objective can be very complex. When we are faced with a problem, to solve this problem is not what I mean by \emph{objective}. Objective means what the agent can gain from solving the problem. Take the previous example of going to university, what's the objective? I can imagine a few:

\begin{quote}
(1) Receive a good education in the field of interest \\
(2) Receive a good general education \\
(3) Have a good job in future \\
(4) Have a good life in future(not only professional, but overall)
\end{quote}

As you can see, there are means and end relationship among the objectives -- for example, \emph{receive a good education in the field of interest} can be seen as a means to \emph{have a good job in the future}. Then how do we fix one objective in order to make a decision? This is actually a very creative process. First, it's reasonable to assume that each agent has some values, or it's impossible for the agent to make any reasonable decisions. The creative process is try to create a network of choices that bridges between the original problem and the values, as illustrated in the following graph:

\begin{center}
  \tikzstyle{line} = [draw, -latex']

  \begin{tikzpicture}[every node/.style={draw, font=\small, text centered, rounded corners, node distance=3cm}, choice/.style={circle, draw}]
    \node  (problem1) {Problem 1};
    \node [choice, fill=red, above right of=problem1] (1a) {1};
    \node [choice, right of=problem1] (1b) {2};
    \node [choice, below right of=problem1] (1c) {3};
    \node [right of=1b] (problem2) {Problem 2};
    \node [choice, above right of=problem2] (2a) {4};
    \node [choice, fill=red, below right of=problem2] (2c) {5};
    \node [below right of=2a] (problem3) {Problem 3};
    \node [choice, right of=problem3] (o3) {Values};

    \path [line] (problem1) -- (1a);
    \path [line] (problem1) -- (1b);
    \path [line] (problem1) -- (1c);

    \path [line] (1a) -- (problem2);
    \path [line] (1b) -- (problem2);
    \path [line] (1c) -- (problem2);

    \path [line] (problem2) -- (2a);
    \path [line] (problem2) -- (2c);

    \path [line] (2a) -- (problem3);
    \path [line] (2c) -- (problem3);

    \path [line] (problem3) -- (o3);
  \end{tikzpicture}
\end{center}

In the graph above, in order to solve the first problem, we generate three options, which in turn becomes a problem itself. To solve the second problem, we generate two options. Now the third problem can be solved using the values of the agent as objective, we suppose the option 5 is chosen. In turn, the option 5 can serve as an objective for problem 2, then the option 1 is chosen -- that's how the first problem is solved. In reality, the graph construction process may begin from both ends in order to bridge the original problem with values of the agent. The intermediate options generated depends on the agent's beliefs, skills, environment, etc. For most problems, there are infinite possible ways on how to bridge between the original problem and the agent's values,  it's a very creative process.

Human beings can profit their linguistic capabilities to generate action plans creatively. The simplest method is to negate a sentence. For example, from \emph{visit Paris} we can get \emph{don’t visit Paris}. Substitution is also a widely used technique. For example, instead of going to visit Paris, one may go to Geneva. The natural language is a huge source of creativity in generating action plans.

Note that the quality of action plans matters more than the quantity. The quality of an action plan is determined by how well it can achieve the specific objective in the concrete scenario. For example, if you want to go from Geneva to Paris in the shortest delay, there are a lot of transportation possibilities, among them one is the best. A person generating the best action plan is more creative than the person who can’t, even though the latter might generate more but lower quality plans. However, in many scenarios, quantity matters as well, as only through comparison we can be sure if a choice is optimal or not. Shopping is a typical example.

According to the new definition, free will is creative -- the better action plans are, more free the agent is. The creativity of agents varies in degrees. A person is more creative than a child or a dog in how to get the banana which they can’t reach. A chess program is usually more creative than a chess master in generating alternative plans, though they generate plans by completely different processes. A poet is more creative than a computer in writing poems.

The definition also implies that free will comes in degrees. In the same scenario with the same objective, the agent who can generate and decide on better action plan is more free than the other one.

This definition implies free will can be improved. If an agent is able to learn to generate better action plans, it will become more free. It’s consistent with our intuitions, and can easily provide an answer to the question \emph{when does a child have free will}. A chess program that can learn from failures to generate better action plans in future will become more and more free. Does this mean free will is a synonym of intelligence? To some extent, I admit the two concepts are closely related. However, as intelligence is an obscure concept to be defined, I will not pursue the topic here.

\subsection{Thematicity}

Free will is not a universal currency that can be used everywhere, it's more like tickets which can be used at specific places.

Talk of free will should be restricted to a proper topic. The correct grammar for talking about free will is \emph{X exhibits free will in T}. The topic T can’t be omitted, or the sentence is incomplete and incomprehensible unless T is implicit in the context. For example, chess program exhibits free will in playing chess games, but not in speaking english. A child exhibits free will in movements of one’s body(mouth, hand, foot, etc), but not in the circulation of blood, digestion or cell fission. However, a doctor may exhibit free will in circulation of blood, digestion and even cell fission by employing his professional knowledge to make plans to tune the functioning of his body. A musician may  exhibit free will in music composition, but a man who doesn’t know music at all can’t.

Free will is not a nature of human beings, but a nurture. The more one learns on a topic, the more creative thus more free he would be on the topic. If one doesn't know anything about a topic, he would not be very different from a cow attending a concert. Millionaires are rarely respected by artists just because they buy and sell masterpieces of art.

For human beings, it's impossible for one to be an expert on all the topics he faces in life. So we delegate health to doctors, social issues to politicians, education of child to schools, etc. However, there is one thing we can't delegate, that's how to spend the whole life. If we just follow the prototypical life roads of the public opinion, we're not showing any free will on the topic of life, even though we are very successful in our professions. To be more free in one's life, we have to learn how to be more creative in life choices.

\subsection{Rationality}

As I argued before, the capacity to decide can't be irrational, or it doesn't qualify the name of free will. However, there’s a weakness here because the rationality of different agents may vary in degrees, far from being perfect. Generally, if we can justify the decision of an agent with respect to the specific objective in the concrete scenario based on the history of the agent, then we can say that the agent decides rationally.

However, generate action plans and choose rationally don’t necessarily imply that free will involves deliberative processes. The process may also be consciousless computations. For example, a chess program can also generate action plans internally and choose the best one in a concrete scenario in order to win the chess.

\subsection{Summary}

In this section, based on the critique of \emph{the myth of choice}, I advocate that it's the capacity to generate choices instead of the capacity to choose that is important to free will. Then I put forward a new definition of free will, which is creative, varies in degrees and can be improved.

In the next section, I'm going to discuss possible attacks on the new conception of free will and respond to them.
